---
title: "Sample Size Estimation for Machine Learning"
author: "Jose Ventura-Leon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sample Size Estimation for Machine Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Determining the minimum sample size needed to achieve a target performance in
Machine Learning models is a fundamental challenge in applied research. Unlike
traditional statistical analysis where tools like G*Power provide closed-form
solutions, machine learning lacks analytical formulas for sample-size planning.

The `ml_sample_size()` function in easyML implements a Monte Carlo simulation
approach that:

1. Generates **learning curves** via repeated model evaluation
2. Fits a **power-law model** for extrapolation
3. Identifies the sample size meeting explicit **decision criteria**

This vignette demonstrates the complete workflow for sample size estimation.

## Quick Start

```{r quickstart, eval=FALSE}
library(easyML)

# Basic usage with simulated data
result <- ml_sample_size(

task = "classification",
  metric = "auc",
  target = 0.80,
  reps = 100
)

# View recommendation
result$recommend_n
```

## Core Concepts

### The Learning Curve

A learning curve plots model performance as a function of training sample size.
It typically shows:

- **Rapid improvement** at small sample sizes
- **Progressively slower gains** as data increases
- **Asymptotic plateau** determined by problem difficulty

### Power-Law Model

The learning curve is modeled as:

$$\text{Performance}(n) = a - b \cdot n^{-c}$$

Where:

- $a$ = asymptotic performance (maximum achievable)
- $b$ = improvement magnitude
- $c$ = convergence rate

### Decision Criteria

The recommended sample size $n^*$ is the smallest $n$ satisfying:

1. **Mean performance** ≥ target
2. **Probability of success** $p_k$ ≥ 0.80 (at least 80% of runs meet target)
3. **Stability** (optional): SD ≤ threshold

## Example 1: Basic Classification

Let's estimate the sample size needed to achieve AUC ≥ 0.80 with Random Forest.

```{r example1, eval=FALSE}
library(easyML)

# Simulate a classification problem
result <- ml_sample_size(
  task = "classification",
  metric = "auc",
  target = 0.80,
  n_grid = c(100, 200, 300, 400, 600, 800, 1000),
  reps = 100,  # 100 Monte Carlo repetitions
  sim_params = list(
    p = 20,           # 20 predictors
    prevalence = 0.20, # 20% positive class
    signal = 2.5       # Moderate signal (AUC ~0.85-0.90 achievable)
  ),
  seed = 123
)
```

### Interpreting the Output

```{r interpret, eval=FALSE}
# View full results
print(result)

# Recommended sample size
result$recommend_n  # e.g., 600

# Summary table with SE of success proportion
result$summary[, c("n", "mean", "sd", "p_meet", "se_p_meet")]
#     n  mean    sd p_meet se_p_meet
# 1 100 0.760 0.025   0.12     0.032
# 2 200 0.795 0.018   0.45     0.050
# 3 300 0.815 0.015   0.78     0.041
# 4 400 0.825 0.012   0.92     0.027
# 5 600 0.840 0.010   0.98     0.014
# ...
```

### Key Output Components

| Component | Description |
|-----------|-------------|
| `mean` | Average metric across R repetitions |
| `sd` | Standard deviation (variability) |
| `p_meet` | Proportion of runs meeting target ($p_k$) |
| `se_p_meet` | Standard error of $p_k$ |

### Visualizing the Learning Curve

```{r plot1, eval=FALSE}
# Plot the learning curve
plot_learning_curve(result)
```

The plot shows:

- Points: observed mean performance
- Shaded band: 95% confidence interval
- Horizontal line: target performance
- Vertical line: recommended $n^*$
- Curve: fitted power law

## Example 2: Choosing the Number of Repetitions (R)

The precision of $p_k$ depends on R. The standard error follows:

$$SE(p_k) \approx \sqrt{\frac{p_k(1-p_k)}{R}}$$

| R | SE at p=0.80 | Use Case |
|---|--------------|----------|
| 50 | 0.057 | Exploratory |
| 100 | 0.040 | Standard |
| 200 | 0.028 | Publication |

```{r reps_guidance, eval=FALSE}
# For exploratory analysis
result_quick <- ml_sample_size(
  task = "classification",
  target = 0.80,
  reps = 50  # Quick exploration
)

# For publication-quality results
result_precise <- ml_sample_size(
  task = "classification",
  target = 0.80,
  reps = 200  # High precision
)
```

The function automatically displays SE guidance:

```
Monte Carlo settings:
  - Repetitions (R): 100
  - Approx SE(p_k) at p=0.80: 0.04
```

## Example 3: Small Datasets (N < 1000)

When your dataset is small, a single test partition may yield unstable estimates.
The **robustness protocol** uses multiple independent partitions.

```{r small_data, eval=FALSE}
# Load a small dataset
data(PimaIndiansDiabetes, package = "mlbench")

# Use 5 independent test partitions
result <- ml_sample_size(
  data = PimaIndiansDiabetes,
  formula = diabetes ~ .,
  task = "classification",
  metric = "auc",
  target = 0.75,
  reps = 100,
  n_outer_splits = 5,  # 5 independent partitions
  verbose = TRUE
)
```

### Interpreting Multiple Partition Results

```{r multi_partition, eval=FALSE}
# View results across partitions
result$outer_splits_results
# $median: 100      # Median n* (use this as recommendation)
# $range: [50, 150] # Range of estimates
# $cv: 0.32         # Coefficient of variation

# If CV > 50%, there's high uncertainty
# Report: "n* = 100 (range: 50-150, CV = 32%)"
```

**Guideline**: If the range exceeds 50% of the median, report the full range
and acknowledge high uncertainty.

## Example 4: Bootstrap Confidence Intervals

For formal inference about $n^*$, use bootstrap CIs:

```{r bootstrap, eval=FALSE}
result <- ml_sample_size(
  task = "classification",
  target = 0.80,
  reps = 100,
  bootstrap_ci = TRUE,  # Enable bootstrap
  n_boot = 1000         # 1000 resamples
)

# View 95% CI
result$recommend_n_ci
# $ci_lower: 400
# $ci_upper: 800
# $ci_median: 600
```

**Interpretation**: "The recommended sample size is n* = 600 (95% CI: 400-800)."

## Example 5: Comparing Multiple Models

Compare sample requirements across algorithms:

```{r multi_model, eval=FALSE}
result <- ml_sample_size(
  task = "classification",
  model = c("rf", "xgboost", "glm"),  # Compare 3 models
  metric = "auc",
  target = 0.80,
  reps = 50
)

# View recommendations by model
result$recommend_n
# $rf:      600
# $xgboost: 600
# $glm:     200

# GLM needs less data but may have lower asymptotic performance
# RF/XGBoost need more data but achieve higher maximum AUC
```

**Trade-off**: Simpler models (GLM) reach targets with less data but have
lower ceilings. Complex models (RF, XGBoost) need more data but achieve
better asymptotic performance.

## Example 6: Checking Fit Quality Before Extrapolation

Always verify the power-law fit before extrapolating:

```{r diagnostics, eval=FALSE}
# Check fit diagnostics
result$curve_diagnostics
# $pseudo_r2: 0.967    # Good: > 0.95
# $rmse_fit: 0.008     # Low fitting error
# $fit_quality: "good" # Automatic classification

# Fit quality interpretation:
# - "good":       R² ≥ 0.95 (safe to extrapolate)
# - "acceptable": R² ≥ 0.80 (extrapolate with caution)
# - "poor":       R² < 0.80 (do not extrapolate)
```

## Example 7: Extrapolation Functions

### Estimating N for a Specific Target

```{r estimate_n, eval=FALSE}
# What n is needed for AUC = 0.85?
estimate_n(result, target_metric = 0.85)
# [1] 1250

# Warnings are issued if:
# - Target exceeds asymptote
# - n exceeds 2x max observed
# - Fit quality is poor
```

### Finding the Plateau

```{r plateau, eval=FALSE}
# Where does performance plateau?

# Absolute delta (within 0.01 of asymptote)
estimate_n_plateau(result, delta = 0.01)
# [1] 8500

# Relative delta (within 1% of target)
estimate_n_plateau(result, delta = 0.01, relative = TRUE)
# Using relative delta: 0.01 x 0.8 = 0.008
# [1] 12000
```

**Interpretation**: Beyond n = 8,500, additional data yields < 0.01 AUC improvement.

## Example 8: Regression Tasks

The method works for regression with appropriate metrics:

```{r regression, eval=FALSE}
result <- ml_sample_size(
  task = "regression",
  metric = "rmse",        # Or "mae", "r2"
  target = 1.0,           # Target RMSE ≤ 1.0
  target_dir = "lower_better",  # Auto-detected for RMSE
  reps = 100,
  sim_params = list(
    p = 15,
    noise_sd = 1.5  # Controls signal-to-noise ratio
  )
)
```

## Best Practices

### 1. Start with Exploration

```{r best1, eval=FALSE}
# Quick exploration with low R
result_explore <- ml_sample_size(
  task = "classification",
  target = 0.80,
  reps = 30,
  n_grid = seq(100, 1000, by = 100)
)

# Refine with higher R after identifying the region
result_final <- ml_sample_size(
  task = "classification",
  target = 0.80,
  reps = 200,
  n_grid = c(300, 400, 500, 600, 700, 800)  # Focus on relevant range
)
```

### 2. Always Check Diagnostics

```{r best2, eval=FALSE}
# Before extrapolating:
if (result$curve_diagnostics$fit_quality == "poor") {
  warning("Do not extrapolate - use only observed values")
}
```

### 3. Report Uncertainty

Good reporting practice:

> "Using Monte Carlo simulation with R = 100 repetitions, we estimated
> n* = 600 (95% CI: 400-800) to achieve AUC ≥ 0.80 with probability
> ≥ 0.80. The power-law fit quality was good (pseudo-R² = 0.97)."

### 4. Consider Hyperparameter Tuning

The $n^*$ estimate assumes fixed hyperparameters. If you plan extensive tuning:

```{r best4, eval=FALSE}
# Add a safety margin (e.g., 20-30%)
n_star <- result$recommend_n
n_with_tuning <- ceiling(n_star * 1.25)
```

## Troubleshooting

### Curve Fit Fails

```{r trouble1, eval=FALSE}
# If curve_fit is NULL or fit_quality is "poor":
# 1. Increase reps
result <- ml_sample_size(..., reps = 200)

# 2. Expand n_grid range
result <- ml_sample_size(..., n_grid = c(50, 100, 200, 500, 1000, 2000, 5000))

# 3. Use only interpolation (don't extrapolate)
```

### High Variability in n*

```{r trouble2, eval=FALSE}
# If CV > 50% with multiple partitions:
# 1. This is expected for small datasets - report the range
# 2. Consider using the upper bound for safety
# 3. Acknowledge uncertainty in your report
```

### Target Seems Unachievable

```{r trouble3, eval=FALSE}
# If recommend_n is NA or Inf:
# 1. Check if target exceeds the asymptote
result$curve_fit  # Look at coefficient 'a'

# 2. Lower your target or increase signal strength
result <- ml_sample_size(..., target = 0.75)  # Lower target

# 3. Use a different algorithm
result <- ml_sample_size(..., model = "glm")  # Simpler model
```

## Summary Table

| Scenario | Recommendation |
|----------|----------------|
| **Exploratory** | R = 50, single partition |
| **Standard** | R = 100, check diagnostics |
| **Publication** | R ≥ 200, bootstrap CI |
| **Small N (<1000)** | n_outer_splits = 5-10 |
| **Extrapolation** | Verify R² > 0.95, n < 2× max observed |
| **With tuning** | Add 20-30% safety margin |

## References
Ventura-Leon, J. (2026). Sample Size Estimation for Machine Learning via
Monte Carlo Simulation: Learning Curves and Power Laws.

Figueroa, R. L., et al. (2012). Predicting sample size required for
classification performance. *BMC Medical Informatics and Decision Making*, 12(1), 8.

## Session Info

```{r sessioninfo, eval=FALSE}
sessionInfo()
```
